# Exercise 6, Week 6: Analysis of longitudinal data


```{r, message=FALSE, warning=FALSE}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", 
                   sep  =" ", header = T)

BPRSL <- read.table("https://raw.githubusercontent.com/tuohmas/IODS-project/master/data/bprsl.csv",
                    sep = ";", header = T)
RATSL <- read.table("https://raw.githubusercontent.com/tuohmas/IODS-project/master/data/ratsl.csv",
                    sep = ";", header = T)

library(tidyverse)

# Turn categorical variables to factors
BPRSL$treatment <- BPRSL$treatment %>% factor()
BPRSL$subject <- BPRSL$subject %>% factor()

RATSL$ID <- RATSL$ID %>% factor()
RATSL$Group <- RATSL$Group %>% factor() 

# Glimpse data
glimpse(BPRSL)
glimpse(RATSL)
```
## Analysis of BPRS data

```{r, message=FALSE, warning=FALSE}
library(ggplot2)

# BPRSL individuals on the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```
To account for "tracking" (individuals who have higher BPRS values at the 
beginning tend to have higher values throughout the study), we will standardize
data by subtrackting every observation by its group mean and dividing that the
corresponding standard deviation.

```{r, message=FALSE, warning=FALSE}
# Create standardized object BPRSL
BPRSL <- BPRSL %>%
  group_by(week) %>%
  mutate(stdbprs = (bprs - mean(bprs))/sd(bprs)) %>%
  ungroup()

# Plot again with the standardised bprs
ggplot(BPRSL, aes(x = week, y = stdbprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")
```
## Mean profiles and mean differences

Let's create mean profiles of our two treatment groups (summarising BPRSL means)
by group weekly).

```{r, message=FALSE, warning=FALSE}
# There are 20 subject per group
n <- 20

# Summarize data with mean and standard error of bprs by treatment and week 
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise(mean = mean(bprs), se = sd(bprs)/sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSS)

# Plot the mean profiles
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.4) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")
```
Standard errors notwhistanding, a clear downward pattern appears for both groups
that was previously unnoticeable. From here alone we cannot clearly determine
which group, if any shows tendency for lower scores: the lines intersersect twice.

A simple boxplot can tell us a better story.

```{r, message=FALSE, warning=FALSE}

# Create a summary data by treatment and subject with mean as the summary var
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>% # ignoring the baseline
  group_by(treatment, subject) %>% 
  summarise(mean=mean(bprs)) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
```

The plot shows very little difference between treatment groups, although group 2
median and 25. percentage points are both slightly lower than is counterpart's.

We also seem to have a single outlier we want to get rid of. Let's investigate 
if this in fact the sole execption, and then filter the data set so to remove it.

```{r, message=FALSE, warning=FALSE}

# Locate outliers
BPRSL8S[rev(order(BPRSL8S$mean)), ] %>% head(n = 5)

# Filter out the single outlier by filtering out means greater than 53.5:
BPRSL8S1 <- BPRSL8S %>% filter(mean < 54)

# See how many observations were removed
nrow(BPRSL8S1) - nrow(BPRSL8S)
```
To investigate possible differences even further in order to attach them with a
formal test result and confidence interval, let's run t-test with the null hypothsis
that the difference in means is zero.

```{r, message=FALSE, warning=FALSE}
# Perform a two-sample t-test
t.test(mean ~ treatment, data = BPRSL8S1, var.equal = TRUE)

```
t-test shows no statistically significant difference between bprs means between 
groups 1 and 2 overall ( _p_ = 0.61). The 95% confidence interval is wide and 
includes zero, driving this point home even harder.

### Analysis of Covariance of the BPRS Data

Baseline measurements of the outcome variable in a longitudinal study are
often correlated with the chosen summary measure and using such measures
in the analysis can often lead to substantial gains in precision when used ap-
propriately as a covariate in an analysis of covariance. 

To calculate this, we might run analysis of variance between groups pre- and 
post-treatment. we can do this with anova() function for fitted model object
(fit), where mean is the target variable and baseline (measurments pre-treatment)
and treatment are explanatory variables.

```{r, message=FALSE, warning=FALSE}
# Add the baseline from the original data as a new variable to the summary data
BPRSL8S2 <- BPRSL8S %>%
  mutate(baseline = BPRS$week0)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + treatment, data = BPRSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

Our two-way ANOVA tests a null-hypothesis that there is no difference in means
pre- and post-treatment. We see that the baseline BPRS is strongly related to the
BPRS values taken after treatment has begun, but there is still no evidence of a
treatment difference even after conditioning on the baseline value.

## Analysis of RATS data

To begin, we will plot the data, identifying the observations in each group but 
ignoring the longitudinal nature of the data (i.e., assume that all the 
observations are independent from one another).

Additionally, we explore whether there is a measurable relationship between body
weight and time and the group by fitting the into linear regression model with
former as a target variable and the two latter as explanatory variables.
```{r, message=FALSE, warning=FALSE}

# plot individual rat's growth profiles
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")

# create a regression model RATS_reg
RATS_reg <- lm(Weight ~ Time + Group, data = RATSL)

# print out a summary of the model
summary(RATS_reg)

```
A clear difference in rat body weight between Group 1 and other two groups. This
difference is also measurable, with very slim chance to be accidental. In addition
the regression on time is also highly significant.

### Fitting Linear Mixed Models to the Rat Data

Taking into account the longitudinal nature of the data, let's fit a regression
model


```{r, message=FALSE, warning=FALSE}
library(lme4)

# Create a random intercept model
RATS_ref <- lmer(Weight ~ Time + Group + (1 | ID), data = RATSL, REML = FALSE)

# Print the summary of the model
summary(RATS_ref)
```

```{r, message=FALSE, warning=FALSE}
# create a random intercept and random slope model
RATS_ref1 <- lmer(Weight ~ Time + Group + (Time | ID), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref1)

# perform an ANOVA test on the two models
anova(RATS_ref1, RATS_ref)
```

```{r, message=FALSE, warning=FALSE}
# create a random intercept and random slope model with the interaction
RATS_ref2 <- lmer(Weight ~ Time + Group + (Time | ID) + (Time * Group), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref2)

# perform an ANOVA test on the two models
anova(RATS_ref2, RATS_ref1)

# draw the plot of RATSL with the observed Weight values
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Observed weight (grams)") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(RATS_ref2)

# Create a new column fitted to RATSL
RATSL <- RATSL %>% mutate(Fitted = Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")
```